A microglia clonal inflammatory disorder in Alzheimer’s Disease

Version 1

Abstract

Summary

Somatic genetic heterogeneity resulting from post-zygotic DNA mutations is widespread in human tissues and can cause diseases, however few studies have investigated its role in neurodegenerative processes such as Alzheimer’s Disease (AD). Here we report the selective enrichment of microglia clones carrying pathogenic variants, that are not present in neuronal, glia/stromal cells, or blood, from patients with AD in comparison to age-matched controls. Notably, microglia-specific AD-associated variants preferentially target the MAPK pathway, including recurrent CBL ring-domain mutations. These variants activate ERK and drive a microglia transcriptional program characterized by a strong neuro-inflammatory response, both in vitro and in patients. Although the natural history of AD-associated microglial clones is difficult to establish in human, microglial expression of a MAPK pathway activating variant was previously shown to cause neurodegeneration in mice, suggesting that AD-associated neuroinflammatory microglial clones may contribute to the neurodegenerative process in patients.

One-Sentence Summary: A subset of Alzheimer Disease patients carry mutant microglia somatic clones which promote neuro-inflammation.

Introduction

Neurodegenerative diseases are a frequent cause of progressive dementia. Alzheimer’s disease (AD) is diagnosed in ∼90% of cases, with an estimated prevalence of ∼10% in the population over 65 years of age 1,2. The role of germline genetic variation in neurodegenerative diseases and AD has been studied intensely. Although autosomal dominant forms of AD due to rare germline variants with high penetrance account only for an estimated ∼1% of cases 3–8, a number of common variants were also shown to contribute to disease risk. Carriers of one germline copy of the epsilon4 (E4) allele of the apolipoprotein E gene (APOE4), present in ∼15 to 20% of population, have a three-fold higher risk of AD, while two copies (∼2 to 3 % of population) increase the risk by ∼10-fold 9–12. Genome-wide association studies (GWAS) have identified an additional ∼50 common germline variants that more moderately increase the risk of AD, including TREM2, CD33, and MS4A6A variants 13–15. Interestingly, the APOE4 allele is responsible for an increased inflammatory and neurotoxic response of microglia and astrocytes in the brain of carriers 16–18, and it was noted that the majority of the other germline AD-risk variants are located within or near genes expressed in microglia 15 and in particular at microglia-specific enhancers 19. These data, together with transcriptional studies 20–22 support the hypothesis that genetic variation in microglia may contribute to the pathogenesis of neurodegeneration and AD.

Somatic genetic heterogeneity (mosaicism), resulting from post-zygotic DNA mutations, is widespread in human tissues, and a cause of tumoral, developmental, and immune diseases 23–26. Aditionally, a role of somatic variants in neuropsychiatric disorders is also suspected 27. Mosaicism has been documented in the brain tissue of AD patients in several deep-sequencing studies 28–30, showing that the enrichment of putative pathogenic somatic mutations in the PI3K-AKT, MAPK, and AMPK pathway do occur in the brain of patients in comparison to controls 30. However these studies performed in whole brain tissue lacked cellular resolution and mechanistic insights, and the role of somatic mutants in neurodegenerative diseases remains poorly understood 23. Somatic variants that activate the PI3K-AKT-mTOR or MAPK pathways in neural progenitors are a cause of cortical dysplasia and epilepsy 31–34 and developmental brain malformations 35, while somatic variants that activate the MAPK pathway in brain endothelial cells are associated with arteriovenous malformations 36. Interestingly, we reported that expression of a somatic variant activating the MAPK pathway in microglia causes neurodegeneration in mice 37, but the presence and contribution of microglial somatic clones in neurodegenerative diseases and AD remains unknown.

Here, we investigated the presence and nature of somatic variants in brain cells from control and AD patients. In an attempt to examine all brain cells at the same resolution, nuclei from neurons, glia cells and microglia, which only represent ∼5% of brain cells, were pre-sorted. In addition, although human microglia are reported to develop in embryo and renew by local proliferation within the brain 38–40, bone marrow-derived myeloid cells can enter the brain, in particular during pathological processes. Therefore we also analyzed matched peripheral blood in order to identify shared somatic mutants between microglia and blood. Finally, in order to achieve high sensitivity in the detection of variants that confer a proliferative or activation advantage (driver mutations) and support the emergence or pathogenicity of mosaic clones 41, and/or that have been previously associated with neurological diseases, we initially performed a targeted deep-sequencing of a panel of 716 genes covering somatic variants reported in clonal proliferative disorders and neurological diseases. We found that microglia from AD patients were enriched for pathogenic variants in comparison to age-matched controls. Furthemore, we found that these microglia-specific AD-associated variants preferentially target the MAPK pathway, including recurrent CBL ring-domain mutations. In addition, we showed that these variants drive a microglia transcriptional program characterized by a strong neuro-inflammatory response previously associated with neurotoxicity, including the production of IL1 and TNF, both in in vitro microglia models and in patients. The natural history of the AD-associated microglia clonal inflammatory disorder we describe here is difficult to establish. Specifically, we do not know whether it contributes to the onset of the neuro-inflammatory process at an early stage of the disease, or if microglia carrying driver mutations preferentially expand later during the course of the disease in response to tissue inflammation. Under both hypotheses however, the presence of neuro-inflammatory microglial clones may contribute to the neurodegenerative process in a subset of AD patients. This report reveals a previously unrecognized presence of AD-associated microglia harboring pathogenic somatic variants in humans and provides mechanistic insight for neurodegenerative diseases by delineating cell-type specific variant recurrence.

Results

Clonal diversity among brain cells and blood from controls and AD patients

We examined post-mortem frozen brain samples and matching blood from 45 patients with intermediate-onset sporadic AD and 44 control individuals who died of other causes, including 27 donors age-matched donors with the AD cohort (Fig. 1A; Table S1). APOE risk allele frequency for patients and controls was comparable to published series 10–12 (Fig. S1A), and analysis of germline mutations did not identify deleterious variants in the 140 genes associated with neurological diseases. Myeloid/microglia, neurons, and glia/stromal cells were purified by flow cytometry using antibody against PU.1 and NeuN 42 (Fig. 1B, Fig. S1B and S1C). Single nuclei (sn)RNAseq was performed on PU.1+ nuclei from one control and 3 AD patients to evaluate microglia enrichment following PU.1+ purification, and a cell-type annotation analysis indicated that ∼94% of PU.1 nuclei correspond to microglia (Fig. 1C and Fig. S1D-S1H). An average of 2.5 brain samples were analyzed for each donor, including cortex samples obtained from all donors, and hippocampus samples, mostly obtained from AD patients (Fig. 1A; Table S1). A total of 744 DNA samples from blood, PU.1+ nuclei, NeuN+ nuclei, and Double Negative nuclei (glia/stromal cells) from patients and controls (Fig. 1A) were submitted to targeted hybridization/capture and deep-DNA targeted sequencing (TDS, Fig. 1D, see Methods), at mean coverage of ∼1,100x (Fig. S1I), for a panel of 716 genes (3.43 Mb, referred to below as BRAIN-PACT) which included genes reported to carry somatic variants in clonal proliferative disorders (n=576 genes) 43,44 or implicated in neurological diseases (n=140) 45–53 (Table S2, see Methods).

Detection of mutations in brain cell types and blood

(A) Table with patient and sample information. (B) Schematic represents the isolation and labelling of nuclei from post-mortem frozen brain samples from controls and Alzheimer’s disease patients with DAPI and antibodies against PU.1+ (myeloid/microglia) and NeuN+ (neurons).Representative flow cytometry dot-plot of nuclei separation. Double negative nuclei are labeled ‘DN’. (C) Percentage of cell types obtained in sorted PU.1+ nuclei determined by single-nuclei RNAseq in 5 brain samples from 4 individuals. (D) Schematic represents the sequencing strategy. Two algorithms (ShearwaterML and Mutect1) were used for variant calling. After annotation, pathogeneity was determined using OncoKb and ClinVar. (E) Venn diagram represents the number of variants and overlap between the ShearwaterML and Mutect1. Numbers in red indicate pathogenic variants (P-SNV). Validation of variants was performed by droplet digital (dd)PCR on pre-amplified DNA when available. (F) Venn diagrams represent the repartition per cell type of the 826 single-nucleotide variations (SNVs) identified in NeuN+: Neurons, PU.1+: microglia, DN: glia, and matching blood. [Numbers] in red indicate pathogenic variants P-SNV

After QC and filtering of germline variants, variant calling using ShearwaterML and a curated Mutect1 analysis identified 826 somatic synonymous and non-synonymous single-nucleotide-variations (SNVs), at an allelic frequency > 0.3% (mean 1.3%) in the 744 samples, corresponding to an overall variant burden of 0.3 mut/Mb (Fig. 1E). Sixty-six SNV were present in more than one sample (Table S3). Droplet digital-PCR performed on pre-amplification DNA for ∼10% of the 760 unique SNV was positive in 90% of cases (Fig.1E; Table S3). After annotation using the OncoKB 54 and ClinVar 55 databases for disease-associated or causative variants (Fig. 2D-2F; Table S3), 96 unique SNV were classified as Pathogenic (P)-SNV. 40% of these P-SNV were tested by droplet digital-PCR and confirmed in 95% of cases (Fig. 1E; Table S3). Positive and negative results in matching brain samples from individual donors were confirmed in 100% of samples at a mean depth of ∼5000x (range 648-23.000x) (Table S3). A venn-diagram analysis of SNVs detected in PU.1+, NeuN+, DN, and blood samples indicated that most (>90%) SNV and P-SNV were cell-type or tissue specific, with ∼ 5% of SNV and ∼ 8% of P-SNV shared between the blood and brain of individual donors (Fig. 1F; Table S3). These data indicate that targeted deep-sequencing of purified nuclei allows to detect clonal mosaic variants with high sensitivity and specificity. In addition, ‘bar-coding’ of clonal variants across tissues suggest that blood clones have a minor contribution to microglia somatic diversity, consistent with its local maintainance and proliferation 38,39

Pathogenic variants are enriched in microglia from AD patients.

(A) Correlation plot represents the mean number of variants per cell type and donor (n=89) (Y axis), as a function of age (X axis). Each dot represents mean value for a donor. Statistics: fitted lines, the correlation coefficients (rs) and associated p values were obtained by linear regression (Spearman’s correlation). (B) Number of SNV per Mb and cell types per donor, of age-matched controls (n=27) and AD patients (n=45). Each dot represents mean value for a donor. Statistics: p-values are calculated with unpaired two-tailed Mann-Whitney U. Note: non-parametric tests were used when data did not follow a normal distribution (D’Agostino-Pearson normality test). (C) Number of SNV per Mb in PU.1 samples across brain regions, of age-matched controls (n=27) and AD patients (n=45). Each dot represents a sample. Statistics: p-values are calculated with Kruskal–Wallis, multiple comparisons. Note: non-parametric tests were used when data did not follow a normal distribution (D’Agostino-Pearson normality test). (D) Correlation plot represents the mean number of pathogenic variants (P-SNV) as determined by ClinVar and/or OncoKB, per cell type and donor (n=89) (Y axis), as a function of age (X axis). Each dot represents mean value for a donor. Statistics: fitted lines, the correlation coefficients (rs) and associated p values were obtained by linear regression (Spearman’s correlation). (E) Number of P-SNV per Mb and cell types per sample, of age-matched controls (n=27) and AD patients (n=45). Each dot represents a sample. Statistics: p-values are calculated with unpaired two-tailed Mann-Whitney U. Note: non-parametric tests were used when data did not follow a normal distribution (D’Agostino-Pearson normality test). (F) Number of P-SNV per Mb in PU.1 samples across brain regions, of age-matched controls (n=27) and AD patients (n=45). Each dot represents a sample. Statistics: p-values are calculated with Kruskal–Wallis test and Dunn’s test for multiple comparis. Note: non-parametric tests were used when data did not follow a normal distribution (D’Agostino-Pearson normality test). (G) Number of P-SNV per Mb and and cell types per donor for age-matched controls (n=27) and AD patients (n=45). Each dot represents mean value for a donor. Statistics: p-values are calculated with unpaired two-tailed Mann-Whitney U test. Odds ratio (95% CI, 2.049 to 29.02) and p values for the association between AD and the presence of driver variants are calculated by multivariate logistic regression, with age and sex as covariates.

Somatic clonal diversity of the different cell types, as evaluated by the SNV/megabase burden was higher in blood (1 mut/Mb) and PU.1+ nuclei (0.5 mut/Mb) than for DN and neurons (0.18 mut/Mb) (Fig. S2A). The SNV/mb burden of blood and PU.1 nuclei increased as a function of age (Fig. 2A; Table S3) as previously reported for proliferating cells 24,56–58. Interestingly, the SNV/mb burden of blood cells from age-matched controls was higher than for AD patients (Fig. 2B). In contrast, there was no difference in SNV/mb burden between PU.1, NEUN and DN samples from AD patients and age-matched controls (Fig. 2B), and between PU.1+ nuclei from the cortex, hippocampus, and brainstem/cerebellum samples (Fig. 2C). These data altogether indicate that the clonal diversity of microglia and blood both increase with age, and that the clonal diversity of blood cells is lower in AD than in age-matched controls who died of other causes including cancer and cardiovascular diseases (see Methods). This is consistent with recent studies showing that clonal hematopoiesis is associated with a higher risk of several diseases related to ageing such as cardiovascular diseases, but inversely associated with the risk of AD 59,60.

Microglia clones carrying pathogenic variants are enriched in AD patients

P-SNV/Mb burden also increased with age (Fig. 2D), however, we found that the P-SNV/Mb burden was selectively and highly enriched in PU.1+ samples from AD patients in comparison to age-matched controls (p=0.0003, Fig. 2E). Analysis of PU1+ P-SNV/Mb burden per brain region indicated that the P-SNV/Mb burden was similar between brain regions within each group (Fig. 2F), and therefore attributable to AD status rather than sampling bias. Moreover, analysis of mutational load per donor confirmed that microglial clones carrying P-SNV were enriched in the brain of AD patients in comparison to age-matched controls (Fig. 2G). Despite the relatively modest cohort size, a logistic regression analysis confirmed the association between the presence of P-SNVs in PU.1 nuclei and AD after adjusting for sex and age (OR= 7; p=0.0035, Fig. 2G and Fig. S2B). In addition, genes targeted by P-SNV were all expressed in microglia (Fig.S2C; Table S3) and the analysis of P-SNV/Mb mutational load restricted to genes that are not expressed in microglia did not show an enrichment of candidate pathogenic variants in AD patients (Fig. S2D; Table S3). Altogether, these results show a cell-specific association between microglia clones carrying P-SNV and AD in this series.

AD patients carry microglial clones with MAP-Kinase pathway variants including recurrent CBL variants

Pathways analysis of genes carrying P-SNV in microglia from AD patients, against the background of the 716 genes sequenced, showed that the most significant pathways enriched were the receptor tyrosine kinase/MAP-Kinase pathways (Reactome, GO, and canonical pathways, Fig. 3A; Table S4), corresponding to driver/oncogenic variants in 6 of the 15 genes of the classical MAPK pathway 61 (CBL, BRAF, RIT1, NF1, PTPN11, KRAS), TEK, and the KEGG Chronic Myeloid Leukemia (CML) pathway, which includes the former plus SMAD5 and TP53 (Fig. 3B, 3C and Fig. S3). Mutational load for MAPK genes was significantly higher in AD patients in comparison to age-matched control (Fig. 3C). Other enriched pathways, albeit less significant, included genes involved in DNA repair and chromatin binding/methyltransferase activity (Fig. 3B; Table S4). No pathway was enriched in age matched controls. P-SNV targeting genes of the classical RTK/MAPK pathway (Fig. 3C) were detected in the PU.1 samples from ∼25% of the AD patients tested (p=0.0145 vs age-matched controls, Fig. 3D and Fig. S3). Strikingly, half of these patients (6 patients, 13 % of AD patients in this series) carried reccurent P-SNV in the RING domain of CBL 62–72 (Fig. 3B-3E). Two additional patients presented with P-SNV in the Switch II domain of RIT1 73 (Fig. 3B-3F). Microglia from the 3 other patients carried activating KRAS (p.A59G), PTPN11 (p.T73I) and TEK (p.R1099*) oncogenic variants previously described in cancer and sporadic venous malformations 74–76 (Fig. 3B and 3D). In addition, a 12th patient carried a gain of function (GOF) U2AF1 (p.S34F) variant 77, which is not a ‘classical MAPK gene’ but activates the MAPK pathway in myeloid malignancies 78 (Fig. 3B and Fig. S3). Two patients carried 2 different MAPK activating variants: microglia from 1 patient carried an activating BRAF (p.L505H) variant 79 in addition to loss of function (LOF) variant CBL (p.C416S), and another patient carried the NF1 (p.L2442*) LOF variant 80,81 in addition to the activating RIT1 (p.M90I) variant (Fig. 3D and Fig. S3). Five patients also carried additional P-SNV targeting genes involved in DNA repair with tumor suppressor function 82,83, including the loss of function variants in ATR (c.6318A>G)84 and SMC1A (p.X285_splice) (Fig. 3D and Fig. S3), and in DNA/histone methylation including TET2 (p.Q1627*) 62,85, IDH2 (p.R140Q) 86, and PBRM1 (c.996-7T>A) 87) (Fig. 3D and Fig. S3). Finally, two patients carried oncogenic variants in genes from the KEGG Chronic Myeloid Leukemia (CML) pathway, SMAD3 (p.R373C) 88 and TP53 (pX261_splice) 89 (Fig. 3D and Fig. S3). The detection of multiple oncogenic variants in the same patients is reminiscent of the features observed in myeloproliferative disorders described outside the brain 72,85.

Somatic microglial clones with multiple and recurrent CBL and MAP-Kinase pathway activating variants.

(A) Pathway enrichment analysis for the genes target of D-SNVs using the panel of 716 genes as background set. Graph shows the most enriched pathways by: Reactome Gene Sets, GO Molecular Functions, Canonical Pathways and KEGG Pathway (see complete list in Table S4). (B) Bar plot indicates the genes carrying D-SNV (y-axis) and the % of AD patients carrying D-SNV for each gene (x-axis). Genes are color-coded by pathway. (C) Representation of the classical MAPK pathway, the 6 genes mutated in AD patients are labeled in red, TEK is labeled in blue, and larger font size indicate reccurence of variants in a given gene. Violin plot shows enrichment in AD patients as compared to age-matched control, p-value: unpaired two-tailed Mann-Whitney U test. (D) Summary Table showing patients carrying D-SNV in the classical RTK/MAPK pathway and CML associated genes (see Table S3) and indicating the detection of variants in blood, and their association with other variants in microglia. (E) Recurrent variants in the ring-like domain of CBL are indicated in red on the diagram structure of gene, and representative western blot from cell lysates from HEK293T cells expressing WT of mutant CBL alleles and stimulated with EGF or control, probed with antibodies against Phospho-p44/42 MAPK (Erk 1/2, Thr202/Tyr204), total p44/42 MAPK (Erk1/2), and HA-tag (BOTTOM). Data are representative from 5 independent experiments. (F) RIT1 M90I and F82L are represented on the 3D structure of the gene (pdb code: 4klz, F82 is within a segment whose structure was not resolved) and representative western blot from HEK293T cells expressing Flag-RIT1 (WT and mutants) and treated -/+ 20% FBS before harvesting. Lysates were probed with antibodies against Phospho-p44/42 MAPK (Erk 1/2, Thr202/Tyr204), total p44/42 MAPK (Erk1/2, (MAPK)), and Flag. Data are representative from 4 independent experiments. (G) Percentage of D-SNVs detected by targeted deep sequencing (TDS) which were also detected by Whole-Exome-Sequencing (WES). (H) Variant allelic frequency (VAF, %) for the BRAFV600E allele in PU.1+ nuclei from brain samples from histiocytosis patients (each dot represents a sample) and for D-SNVs in in PU.1+ nuclei from brain of AD patients (each dot represent a variant). Note: non-parametric tests were used when data did not follow a normal distribution (D’Agostino-Pearson normality test).

Recurrent CBL and RIT1 variants activate the MAPK pathway

CBL is an E3 ubiquitin-protein ligase that negatively regulates RTK signaling via MAPK90. CBL somatic and germ-line LOF variants such as R420Q have been previously associated with tumoral diseases including clonal myeloproliferative disorders 62–72 and RASopathies 91 respectively. We confirmed that CBL RING-domain variants found in AD patients increased MAPK phosphorylation in response to EGF upon expression of HA-tagged WT or mutant alleles in HEK293T cells (Fig. 3E and Fig. S4A). RIT1 is a RAS GTPase, and somatic or germ-line GOF variants such as RIT1 F82L and RIT1 M90I, also enhance MAPK signaling in malignancies 73 and RASopathies 92,93 respectively. As in the case of CBL variants, the 2 RIT1 variants found in AD patients increased MAPK phosphorylation in response to FBS in HEK293T cells expressing these mutant alleles (Fig. 3F and Fig. S4B, S4C). These data altogether indicatesd that a subset of AD patients (12/45, ∼ 27% of this series) present with microglial clones carrying one or several oncogenic variants that activate the RTK/MAPK pathway, and are characterized by recurrent oncogenic variants in CBL and RIT1.

Allelic frequency of the patients’ MAPK activating variants

The allelic frequencies at which MAPK activating variants are detected in brain samples from AD patients range from ∼1 to 6% of microglia (Fig. 3G), which correspond to clones representing 2 to 12% of mutant microglia in these samples, assuming heterozygosity. This allelic frequency is in the range of the allelic frequency of the MAPK activating BRAFV600E variant in microglia from 6 patients diagnosed with BRAFV600E+ histiocytosis, a rare clonal myeloid disorder associated with neurodegeneration 37,94–97 (Fig. 3G; Table S5). These data suggested that the size of the mutant microglial clones in AD patients was compatible with a role in a neuro-inflammatory/neurodegeneration process.

Other variants found in microglia from AD patients

Driver variants that did not involve the MAPK pathway included LOF variants in the DNA repair gene CHEK2 including CHEK2 c.319+1G>A 98 and CHEK2 R346H (Fig. S3 and Fig. S4D), Mediator Complex gene MED12 99, Histone methyltransferases SETD2 100 and KMT2C/MLL3, the DNA methyltransferase DNMT3A 85,101, DNA demethylating enzymes TET2 and the Polycomb proteins ASXL1 85. Of note, TET2, DNMT3 and KMT2C variants when present, were frequently detectable in the patients’ matching blood at low allelic frequency (Fig. S3). TET2, DNMT3 and KMT2C are frequently mutated in clonal hematopoiesis 57,58, suggesting that in contrast to other variants, the presence of TET2, DNMT3 and KMT2C/MLL3 in the brain of patients may reflect the entry of blood clones in the brain.

In half of the AD patients, no microglia driver variants were identified. Targeted deep sequencing (TDS) cannot identify variants located outside of the BRAIN-PACT panel, such as other potential additional variants that would activate the MAPK pathway. Therefore, we performed whole exome sequencing (WES) of PU.1 nuclei at an average depth ∼400x, in selected samples from 48 donors, including samples from most of the patients negative for driver variants by TDS (n=17 out of 22), a selection of patients with variants identified by TDS (n=16 out of 23), and 15 controls, followed by a curated Mutect analysis. Only 6/15 (40%) of the driver SNVs previously identified by TDS and confirmed by ddPCR were detectable by WES in these samples (Fig. 3H), indicating a lower sensitivity of WES. Nevertheless, after annotation by 4 modeling predictors (Polyphen, SIFT, CADD/MSC and FATHMM-XF 102–107 additional SNVs predicted to be deleterious with high confidence were identified in 8/22 patients without driver variants identified by TDS (Fig. S3; Table S6). Interestingly, 4 of the predicted deleterious variants identified by WES targeted genes that regulate the MAPK pathway (ARHGAP9, ARHGEF26, CHD8, and DIXDC1 (Fig. S3; Table S6).

The patients’ MAPK activating variants increases ERK phosphorylation, proliferation, inflammatory and mTOR pathways in murine microglia and macrophages

CBL variants increased ERK phosphorylation upon lentiviral transduction in BV2 murine microglial cells 108,109 (Fig. S4E). However as this line was immortalized by v-Raf, which might interfere with the study of the MAPK pathway, we also stably expressed WT and variant CBL, RIT1, KRAS, PTPN11 alleles in SV-U19–5 transformed mouse ‘MAC’ lines 110,111 (see Methods and Fig. S5A,B). MAC lines expressing CBL, RIT1, KRAS and PTPN11 variants presented with increased ERK phosphorylation and/or increased proliferation in comparison to their WT controls, as measured by Western immunoblotting and EdU incorporation (Fig. 4A and Fig. S5A,B). In addition, Hallmark and KEGG pathway analysis of RNAseq data from control and mutant lines showed increased RAS, TNF, IL6 and JAK STAT signaling, complement, inflammatory responses, and mTOR pathway activation signatures in mutants (Fig. 4B; Table S7). These data indicated that microglia variants from patient’s activate murine microglial cells and growth factor-dependent macrophages with proliferative and inflammatory responses in vitro. However, overexpression of mutant alleles in mouse cell lines does not necessarily recapitulate or predict the effects of a heterozygous genetic variant in physiological conditions. Thus, we investigated the role of CBLC404Y allele in heterozygous human primary microglia-like cells.

MAPK pathway activating variants in mouse macrophages and human iPSC-derived microglia-like cells.

(A) Representative western-blot analysis (Top panels) and quantification (Middle panels) of phospho- and total-ERK in lysates from a murine CSF-1 dependent macrophage cell line expressing CBLWT, CBLI383M, CBLC384Y, CBLC404Y, CBLC416S (n=3-6), and RIT1WT, RIT1F82L and RIT1M90I (n=3), KRASWT and KRASA59G (n=3). Bottom panels depicts flow cytometry analysis of EdU incorporation in the same lines. Statistics, Unpaired t-test. (B) HALLMARK and KEGG pathways (FDR/adj.p value <0.25, selected from Table S7) enriched in gene set enrichment analysis (GSEA) of RNAseq from mutant CSF-1 dependent macrophages lines CBLI383M, CBLC384Y, CBLC404Y, CBLC416S, CBLR420Q, RIT1F82L RIT1M90I, KRASA59G, and PTPN11T73I (n=3-6) in comparison with their wt controls. NES: normalized enrichment score. (C) Sanger sequencing of 2 independent hiPSC clones (#93 and #91) of CBL404C/Y heterozygous mutant carrying the c.1211G/A transition on one allele and 2 independent isogenic control CBL404C/C clones (#71 and #89) all obtained by prime editing. (D) Photomicrographs in CBL404C/C and CBL404C/Y iPSC-derived microglia-like cells.(E) Quantification of leading edge and lateral lamellipodia in CBL404C/C and CBL404C/Y iPSC-derived microglia-like cells. n=3-7, statistics: pvalue are obtained by nested one-way ANOVA. (F) Flow cytometry analysis of cell size for the same lines (n>3) statistics: pvalue are obtained with nested one-way ANOVA,). (G) Flow cytometry analysis of EdU incorporation in CBL404C/C and CBL404C/Y microglia-like cells after a 2 hours EdU pulse. n=3, unpaired t-test). (H) Western-blot analysis (left) and quantification (right) of phospho- and total-ERK proteins in lysates from CBL404C/C and CBL404C/Y microglia-like cells starved of CSF-1 for 4 h and stimulated with CSF-1 (5 min, 100 ng/mL) (n=4), statistics: pvalue are obtained with two-way ANOVA.

Heterozygosity for a CBL variant allele activates human microglia-like cells

We used prime editing 112 of human induced pluripotent stem cells (hiPSCs, see Methods) to generate isogenic hiPSCs clones heterozygous for the patients’ variants (Fig. 3C and Fig. S5). We focused our analysis on CBL404C/Y mutant lines because CBL was mutated in 6 patients and 2 of them carried the same CBL c.1211G>A p.C404Y variant (Fig. 3D). Microglia-like cells were differentiated from two independent hiPSC-derived CBL404C/Y lines and their isogenic CBL404C/C controls (Fig. 4C and Fig. S5C). CBL404C/Y and isogenic CBL404C/C microglia-like cells expressed similar amount of CBL total mRNA and protein, and CBL404C/Y cells expressed wt and mutant mRNA in similar amounts, as expected assuming bi-allelic expression of CBL (Fig. S5D-S5F). CBL404C/Y cells presented with a phenotype comparable to isogenic CBL404C/C microglia-like cells for expression of IBA1, CSF1R, NGFR, EGFR, CD11b, MRC1, CD36, CD11c, Tim4, CD45, and MHC Class II (Fig. S5G). Their viability was also comparable to control (Fig. S5H). However, CBL404C/Y cells were larger and presented with more lamellipodia, resulting in an amoeboid morphology less frequently observed in isogenic controls (Fig. 4D,4E), and their proliferation rate was slightly increased, as measured by EDU incorporation (Fig. 4F). Moreover CBL404C/Y cells cultured in CSF1-supplemented medium also presented with a higher basal pERK level than control when restimulated with CSF1 (Fig. S5I), and ERK phosphorylation after stimulation of starved microglia-like cells with CSF-1 was increased by ∼2 fold in comparison to isogenic WT (Fig. 4G). Altogether, these results showed that heterozygosity for a CBLC404Y allele is sufficient to activate human microglia-like cells increasing their proliferation and ERK activation.

Heterozygosity for a CBLC404Y allele drives a microglial neuroinflammatory/AD associated signature

Gene Set Enrichment Analyses (GSEA) of RNAseq comparing CBL404C/Y and isogenic CBL404C/C microglia-like cells showed upregulation of Glycolysis, Oxidative Phosphorylation, and mTORC1 signatures, indicating increased metabolism and energy consumption by the mutant cells (Fig. 5A; Table S8). In addition, as observed in MAC lines, CBL404C/Y cells upregulated complement, TNF, and JAK STAT signaling and inflammatory signatures (Fig. 5A; Table S8) 113. Increased production of TNF, IL-6, IFN-ψ, IL-1β, C3 and complement Factor H (CFH) by CBL404C/Y cells was confirmed by ELISA (Fig. 5B). In addition, CBL404C/Y microglia-like cells also presented with signatures from the KEGG database associated with neurodegenerative disorders (Fig. 5A; Table S8), and for the recently published human microglia AD scRNAseq signature, obtained by analysis of 24 sporadic AD patients and 24 controls 21 (Fig. 5C). These data indicated that heterozygosity for the CBLC404Y allele is sufficient to drive expression of a neuroinflammatory/AD signature in a human microglia-like cell type, characterized by increased metabolism and the production of neurotoxic cytokines known to interfere with normal brain homeostasis.

CBL404C/Y microglia signature.

(A) HALLMARK and KEGG pathways (FDR/adj.p value <0.25, selected from Table S8) enriched in gene set enrichment analysis (GSEA) of RNAseq from from CBL404C/Y iPSC-derived macrophages and isogenic controls NES, normalized enrichment score. (B) ELISA for pro-inflammatory cytokines (n=3) and complement proteins (n=2) in the supernatant from CBL404C/Y iPSC-derived microglial like cells and isogenic controls. Statistics: p-value are obtained by nonparametric Mann-Whitney U test,* 0.05, ** 0.01, *** 0.001, **** 0.0001. (C) GSEA analysis for enrichment of the human AD-microglia snRNAseq signature (MIC1) 21 in differentially expressed genes between CBL404Y/C microglial like cells and isogenic controls. (D) Unsupervised Louvain clustering of snRNAseq data from 5 samples of FACS-purified PU.1+ microglia nuclei from 4 donors (see Fig. S1C) control C14, AD without driver variant (AD34) and AD with driver variants (AD 52 and 53). (E) Dot plot represents the GSEA analysis of HALLMARK and KEGG pathways enriched in snRNAseq microglia clusters (samples from all donors). Genes are pre-ranked per cluster using differential expression analysis with SCANPY and the Wilcoxon rank-sum method. Statistical analyses were performed using the fgseaMultilevel function in fgsea R package for HALLMARK and KEGG pathways. Selected gene-sets with p-value < 0.05 and adjusted p-value < 0.25 are visualized using ggpubr and ggplot2 R package (gene sets/pathways are selected from fig S6B, Table S9). (F) Dot plot represents the GSEA analysis (as in (E)) of HALLMARK and KEGG pathways enriched in cluster 2/2B and deconvoluted by donor samples (selected from Fig. S6A).

The MAPK variant neuroinflammatory microglial signature is detectable in patients

Unsupervised Louvain clustering of the snRNAseq data from 5 samples of purified microglia nuclei from 4 donors (control, AD without and with driver variants, see Fig. S1C; Table S9) outlined 17 microglia clusters (Fig. 5D, Fig. S1A-E and S7A). GSEA analysis of the snRNAseq data showed that microglia samples from mutant patients were enriched for the signatures observed in the MAC lines and CBL404C/Y cells (Fig. S6). In particular microglia cluster 2/2B, where KRAS A59G variant reads from patient AD52 are detected despite the small size of the mutant clones and the low sensitivity of scRNAseq to detect low allelic frequency variants, was most enriched for the inflammatory, TNF, mTOR and oxidative phosphorylation and glycolysis signatures (Fig. 5E and Fig. S6). Deconvolution of cluster 2/2B by samples confirmed that the samples from patients carrying variants (AD52, AD53) and not the controls (C11, AD34) were responsible for this neuroinflammatory and metabolic signature (Fig. 5F and Fig. S6). Altogether, the above results strongly support the hypothesis that microglial clones with driver/oncogenic variants that activate the MAPK pathway are responsible for a metabolic and neuroinflammatory signature in vitro which includes the production of neurotoxic cytokines that are also detected in vivo in patients.

Discussion

We report here that microglia from a cohort of 45 patients with intermediate-onset sporadic AD (mean age 65 y.o), is enriched for clones carrying driver/oncogenic variants predominantly in the MAPK pathway. These variants are absent from blood, glia or neurons in most cases. They include reccurent MAPK pathway variants (CBL RING domain variants in 6 patients), which promote microglial proliferation, activation, and expression of a neuroinflammatory/neurodegereration-associated transcriptional programme in vitro and in vivo, and the production of neurotoxic cytokines IL1b, TNF, and IFNg 114–117. Heterozygous expression of pathogenic CBL variant in human microglia-like cells was sufficient to drive a transcriptional program that associates with increased metabolic activity and a neurotoxic inflammatory response, also observed in microglia from patients with MAPK-activating variants.

The association between AD and MAPK pathway variants is consistent with a previous study where WES performed on unseparated brain tissue from AD patients showed that putative pathogenic somatic variants were enriched for the MAPK pathway, despite the lower sensitivity of the approach and the lack of cellular specificity 30. The pathogenic role of the somatic pathogenic variants in the MAPK pathway associated with the microglia of AD patients is supported by several lines of evidence. We show here that they promote a neuroinflammatory/neurodegereration-associated transcriptional programme in microglia like cells. In addition, somatic variants that activate the MAPK pathway in tissue macrophages cause a clonal proliferative and inflammatory disease called Histiocytosis, strongly associated with neurodegeneration 37,94–96, and introduction in mouse microglia of the variant allele most frequently associated with histiocytosis (BRAFV600E) causes neurodegeneration in mice 37. The allelic frequencies of pathogenic variants found in AD patients is lower than values classically observed in solid tumors or leukemia, but within the range of the clonal frequency of pathogenic T cells observed in auto-immune diseases 118, and we found that they were in the range of the allelic frequencies observed for the BRAFV600E variant in microglia in the brain of Histiocytosis patients. Moreover, the RAS/MAPK signaling pathway is involved in microglia proliferation, activation and inflammatory response 119–121, neuronal death, neurodegeneration, and AD pathogenesis 15,19,37,122, and its activation has been proposed to be an early event in the pathophysiology of AD in human 123. Neuroinflammation is an early event in AD pathogenesis, increasingly considered as critical in pathogenesis initiation and progression 16,124,125. This is underscored by the observation that the main known genetic risk factor for sporadic AD is the APOE4 allele, responsible for an increased inflammatory response in the brain of APOE4 carriers 16. In this regard, the contributing role of MAPK activating variants could be comparable to that of the APOE4 allele, and we noted that the allelic frequency of APOE4 allele is lower in patients with pathogenic variants (16/46 alleles, 34%) than patients without detected variant (23/44 alleles, 53%) although the difference did not reach significance in this series.

Variants targeting the DNA-repair and DNA/histone methylation pathways are also enriched among AD patients, sometimes associated in the same patients, albeit their functional significance was not investigated here. Of note however, germline variants of the DNA-repair transcription factor TP53, and DNA damage sensors ATR and CHEK2 were shown to promote accelerated neurodegeneration in human 82,83.

Microglia variants are frequently absent from blood, and our DNA sequencing barcoding approach does not support a model where blood cells massively infiltrate the brain or replace the microglia pool in patients from our series, but instead consistent with the local maintenance and proliferation of microglia 38,39. In addition, our results are consistent with a recent study showing that clonal hematopoiesis was inversely associated with the risk of AD60.

The association of microglia clones carrying pathogenic variants with AD in a subset of patients, together with evidence that they drive neuroinflammation, suggest that these clones could contribute to AD pathogenesis, together with other genetic and environmental factors. Lewis bodies, amyloid angiopathy, tauopathy, or alpha synucleinopathy, were equally distributed among AD patients with or without microglia clones carrying MAPK activating variants. The natural history of the microglial clones is difficult to study in human. It is possible that microglial clones with proliferative and activation advantage and a neuroinflammatory and neurotoxic profile may be present at the onset and contribute to the early stages of the disease. Alternatively it is also possible that the microglial clones carrying the driver mutations appear or are selected later during the course of the disease in the inflammatory milieu of the AD brain. In the latter case, microglial clones would not contribute to disease initiation, but may contribute to the progression of neuroinflammation and neurodegeneration.

Competing interests

FG has been a paid consultant (no equity) to Third Rock Ventures from 2018 to 2020. Sequencing costs and analysis in this study were covered in part by a SRA between Third Rock venture and MSKCC. This work led to patents PCT/US2022/037893/WO2023004054A1 ‘Methods and compositions for the treatment of alzheimer’s disease’ by MSKCC and PCT/US2018/047964 ‘Kinase mutation-associated neurodegenerative disorders’ by MSKCC.

Supporting information

eLife Assessment

This is a fundamental study that advances our understanding of the contribution of somatic variations in microglia that may contribute to the onset or progression of neurodegenerative disease. Specifically, during Alzheimer's disease, somatic mutations were identified in the MAPK pathway genes. The findings presented here are backed by compelling evidence drawn from a patient cohort, along with mechanistic proof-of-concept studies. Collectively, this research will be of interest to a wide audience, particularly those involved in the study of somatic mutations, neurodegeneration, immunology, and cell signalling.

Reviewer #1 (Public Review):

In the manuscript "A microglia clonal inflammatory disorder in Alzheimer's Disease", Vicario et al. provide a compelling study elucidating a potential contribution of somatic mutations within the microglia population of the CNS that accelerates microglia activation and disease-associated gene signatures in Alzheimer's disease. Here they especially identified an "enrichment" of pathological SNVs in microglia, but not the peripheral blood, that are associated with clonal proliferative disorders and neurological diseases in a subset of patients with AD. Convincingly, they identified P-SNVs in microglia of AD patients located within the ring domain of CBL, a negative regulator of MAPK signaling. They further provide mechanistic insights into how these variants result in MAPK over-activation and subsequently in a pro-inflammatory phenotype in human microglia-like cells in vitro.

Overall, this study provides clear and detailed evidence from an AD patient cohort pointing to a potential contribution of microglia-specific somatic mutations to disease onset and/or progression in a subset of patients with Alzheimer's disease.

Strengths:

As outlined above, the study identified P-SNVs in microglia of AD patients associated with clonal proliferative disorders, but also gave an in-depth analysis of re-occurring P-SNVs located within the ring domain of CBL, a negative regulator of MAPK signaling. They further provide mechanistic insights into how these variants result in MAPK over-activation and subsequently in a pro-inflammatory phenotype in HEK cells, BV2 cells, MAC cells, and human microglia-like cells in vitro.

Great care was taken here to validate their hypotheses at each step, as well as to identify the limitations of the possible conclusions. For example, they highlight that the pathway proposed to be affected may be an explanation for a subset of AD patients, and emphasize that it is yet unclear whether this accumulation of pathological SNVs is a cause or consequence of disease progression

The study clearly supports an enrichment of P-SNVs in several genes associated with clonal proliferative disorders in microglia and nicely separates this from SNVs associated with clonal hematopoiesis in the peripheral blood found in AD patients and controls.

The authors further acknowledged that several age-matched control patients were diagnosed with cancer or tumor-associated diseases and carefully dissected the occurring SNVs in these patients are not associated with the P-SNVs identified in the microglial compartment of the AD cohort.

Weaknesses:

Even though the study is overall very convincing, several points could help to connect the seen somatic variants in microglia more with a potential role in disease progression. The connection of P-SNVs in the genes chosen from neurological disorders was not further highlighted by the authors.

The authors show in snRNA-seq data that a disease-associated microglia state seems to be enriched in patients with somatic variants in the CBL ring domain, however, this analysis could be deepened. For example, how this knowledge may translate to patient benefits when the relevant cell populations appear concentrated in a single patient sample (Figure 5; AD52) is unclear; increasing the analyzed patient pool for Figure 5 and showcasing the presence of this microglia state of interest in a few more patients with driving mutations for CBL or other MAPK pathway associated mutations would lend their hypotheses further credibility.

A potential connection between P-SNVs in microglia and disease pathology and symptoms was not further explored by the authors.

A recent preprint (Huang et al., 2024) connected the occurrence of somatic variants in genes associated with clonal hematopoiesis in microglia in a large cohort of AD patients, this study is not further discussed or compared to the data in this manuscript.

Reviewer #2 (Public Review):

Summary:

In this study, Vicaro et al. aimed to quantify and characterize mosaic mutations in human sporadic Alzheimer's disease (AD) brain samples. They focused on three broad classes of brain cells, neurons that express the marker NeuN, microglia that express the marker PU.1, and double-negative cells that presumably comprise all other brain cell types, including astrocytes, oligodendrocytes, oligodendrocyte progenitor cells, and endothelial cells. The authors find an enrichment of potentially pathogenic somatic mutations in AD microglia compared to controls, with MAPK pathway genes being particularly enriched for somatic mutations in those cells. The authors report a striking enrichment for mutations in the gene CBL and use in vitro functional assays to show that these mutations indeed induce MAPK pathway activation.

The current state of the AD and somatic mutation fields puts this work into context. First, AD is a devastating disease whose prevalence is only increasing as the population of the U.S. is aging, necessitating the investigation of novel features of AD to identify new therapeutic opportunities. Second, microglia have recently come into focus as important players in AD pathogenesis. Many AD risk genes are selectively expressed in microglia, and microglia from AD brain samples show a distinct transcriptional profile indicating an inflammatory phenotype. The authors' previous work shows that a genetic mouse model of mosaic BRAF activation in macrophages (including microglia) displays a neurodegenerative phenotype similar to AD (Mass et al., 2017, doi:10.1038/nature23672). Third, new technological developments have allowed for identifying mosaic mutations present in only a small fraction of or even single cells. Together, these data form a rationale for studying mosaic mutations in microglia in AD. In light of the authors' findings regarding MAPK pathway gene somatic mutations, it is also important to note that MAPK has previously been implicated in AD neuroinflammation in the literature.

Strengths:

The study demonstrated several strengths.

Firstly, the authors used two methods to identify mosaic mutations:

(1) deep (~1,100x) DNA sequencing of a targeted panel of 716 genes they hypothesized might, if mutated somatically, play a role in AD, and

(2) deep (400x) whole-exome sequencing (WES) to identify clonal mosaics outside of those 716 genes.

A second strength is the agreement between these experiments, where WES found many variants identified in the panel experiment, and both experiments revealed somatic mutations in MAPK pathway genes.

Third, the authors demonstrated in several in vitro systems that many mutations they identified in MAPK genes activate MAPK signaling. Finally, the authors showed that in some human brain samples, single-cell gene expression analysis revealed that cells bearing a mosaic MAPK pathway mutation displayed dysregulated inflammatory signaling and dysregulation in other pathways. This single-cell analysis was in agreement with their in vitro analyses.

Weaknesses:

The study also showed some weaknesses. The sample size (45 AD donors and 44 controls) is small, reflected in the relatively modest effect sizes and p-values observed. This weakness is partially ameliorated by the authors' extensive molecular and functional validation of mutation candidates. Another weakness is the lack of discussion of whether the genes found to be mutated somatically in AD show any AD-risk alleles in the population. If they did, it would further support the authors' conclusions that they are playing a role in AD. Finally, as the authors point out, this study cannot conclude whether microglial mosaic mutations cause AD or are an effect of AD. Future studies may shed more light on this important question.

Conclusions and Impact:

Considering the study's aims, strengths, and weaknesses, I conclude that the authors achieved their goal of characterizing the role of mosaic mutations in human AD. Their data strongly suggest that mosaic MAPK mutations in microglia are associated with AD. The impacts of this study remain to be seen, but they could include attempts to target CBL or other mutated genes in the treatment of AD. This work also suggests a similar approach to identifying potentially causative somatic mutations in other neurodegenerative diseases.

Author response:

The following is the authors’ response to the original reviews.

We edited the manuscript for clarity, added information described in new figure panels (below) and corrected typos.

In figure 1 we corrected a typo.

In figure 2, panel 2H, and Figure S2E, we included a new statistical analysis (mixed effect linear regression) to compare mutational burden in controls and AD patients.

In figure 3, and Figure S4B, we revised the western blots panels in Panel 3E,F, to improve presentation of controls and quantification.

we corrected typos.

In figure 5 we removed a panel (former 5D) which did not add useful information.

In Figure S1A we included information about sex and age from the control and patients analyzed. In Figure S2B, we added an analysis of the mutational burden in controls, distinguishing controls with and without cancer.

We modified Table S1 for completeness of information for all samples analyzed.

Reviewer #1:

Weaknesses:

Even though the study is overall very convincing, several points could help to connect the seen somatic variants in microglia more with a potential role in disease progression. The connection of P-SNVs in the genes chosen from neurological disorders was not further highlighted by the authors.

All P-SNVs are reported in Table S3.

We observed only two P-SNVs within genes associated to neurological disorders (brain panel in Table S2). - SQSTM1 (p.P392L) was identified in blood but not in brain from the patient AD48A.

- OPTN was identified (p.Q467P) in PU.1 from control 25.

To highlight this point, we modified the first paragraph of the discussion as follow:

“We report here that microglia from a cohort of 45 AD patients with intermediate-onset sporadic AD (mean age 65 y.o) is enriched for clones carrying pathogenic/oncogenic variants in genes associated with clonal proliferative disorders (Supplementary Table 2) in comparison to 44 controls. Of note we did not observe microglia P-SNVs within genes reported to be associated with neurological disorders (Supplementary Table 2) in patients, and one such variant was identified in a control (Supplementary Table 3) “.

The authors show in snRNA-seq data that a disease-associated microglia state seems to be enriched in patients with somatic variants in the CBL ring domain, however, this analysis could be deepened. For example, how this knowledge may translate to patient benefits when the relevant cell populations appear concentrated in a single patient sample (Figure 5; AD52) is unclear; increasing the analyzed patient pool for Figure 5 and showcasing the presence of this microglia state of interest in a few more patients with driving mutations for CBL or other MAPK pathway associated mutations would lend their hypotheses further credibility.

We acknowledge this limitation, but we respectfully submit that the analysis was performed in 2 patients. AD 53 also show a MAPK-associated inflammatory signature in the microglia clusters associated with mutations.

We performed the analysis on all FACS-purified PU.1+ nuclei samples that passed QC for single nuclei RNAseq. It should be noted that this analysis is extremely difficult with current technologies because microglia nuclei need to be fixed for PU.1 staining and FACS purification and the clones are small (~1% of microglia).

A potential connection between P-SNVs in microglia and disease pathology and symptoms was not further explored by the authors.

At the population level, Braak/CERAD scores, the presence of Lewy bodies, amyloid angiopathy, tauopathy, or alpha synucleinopathy were not different between AD patients with or without pathogenic microglial clones (Figure S3 and Table S1). Of note, we studied here a homogenous population of AD patients.

At the tissue level, the roles of mutant microglia in plaques for example is being investigated, but we do not have results to present at this time.

A recent preprint (Huang et al., 2024) connected the occurrence of somatic variants in genes associated with clonal hematopoiesis in microglia in a large cohort of AD patients, this study is not further discussed or compared to the data in this manuscript.

This pre-print supports the high frequency of detection of oncogenic variants associated with clonal proliferative disorders, they hypothesize that the mutations may be associated with microglia, but they only check a few mutations in purified microglia. Most of the study is performed in whole brain tissue. It does not really bring new information as compared to other study we cite in the introduction (and to our manuscript).

Reviewer #2 (Recommendations For The Authors):

Suggestions for improved or additional experiments, data, or analyses:

The authors can demonstrate that identified pathological SNVs from their AD cohort also lead to the activation of human microglia-like cells in vitro, but do not provide any data from histological examination of the patient cohort (e.g. accumulation at the plaque site, microglia distribution, and cell number). The study could be further supported by providing a histological examination of patients with and without P-SNVs to identify if microglia response to pathology, microglia accumulation, or phagocytic capacity are altered in these patients.

We performed IBA1 staining in brain samples from control and from AD patients, with or without microglial clones and microglia density was not different between patient with and without mutations. In addition, histological reports from the brain bank (Braak/CERAD scores, Lewis bodies, amyloid angiopathy, tauopathy, or alpha synucleinopathy did not suggest differences between patient with and without mutations (Figure S3). These results are preliminary and further investigations are ongoing.

It would have been interesting to see if for example, transgenic AD mice with an introduced somatic mutation in microglia show an altered disease progression with alterations in amyloid pathology or cognition.

We agree with the reviewer. We performed an in vivo study with mice expressing a 5xFAD transgene, an inducible microglia Cx3cr1CreERt2 BrafLSL-V600E transgene, or both, and performed survival, behavioral (Y-Maze and Novel Object Recognition), and histological analyses for β-Amyloid, p-Tau and Iba1 staining.

Microgliosis was increased in the group with the 2 transgenes, however the phenotype associated with the expression of a BrafV600E allele in microglia (Mass et al Nature 2017) was strongly dominant over the phenotype of 5xFAD mice, which did not allow us to conclude on survival and behavioral analyses.

Other studies with different transgenes are in progress but we have no results yet to include in this revised manuscript.

To connect the somatic mutations in microglia better to a potential contribution in neurodegeneration or neurotoxicity, the authors could provide further details on how to demonstrate if human microglia-like cells respond differentially to amyloid or induce neurotoxicity in a co-culture or slice culture model.

These studies are undertaken in the laboratory, but unfortunately, we have no results as yet to include in this revised manuscript.

The number of samples analyzed for hippocampi, especially in the age-matched controls might be underpowered.

Unfortunately, despite our best efforts, we were not able to analyze more hippocampus from control individuals. To control for bias in sampling as well as to other potential bias in our analysis, we investigated the statistical analysis of the cohorts for inclusion of age as a criterion (age matched controls), inclusion of a random effect structure, and possible confounding factor such as sex, brain bank site, and samples’ anatomical location (see revised Methods and revised Fig. 2C, F, and H, and S2B).

We first tested whether the inclusion of age is appropriate in a fixed-effects linear regression using a generalized linear model (GLM) with gaussian distribution. Compared to the baseline model, the model with age had significantly low AIC (from -66.6 to -71.9, P = 0.0067 by chi-square test). Therefore, the inclusion of age as a fixed effect is appropriate. We next tested multiple structures of mixed-effects linear modeling. We used donors as random effects, while utilizing age, disease status (neurotypical control vs. AD), or both as fixed effects. Fitting was performed using the lme function implemented in the nlme package with the maximum likelihood (ML) method. The incorporation of age and disease status significantly improved overall model fitting. Both age and AD are associated with a significant increase in SNV burden in this model (P<1x10^-4 and P=1x10^-4, respectively, by likelihood ratio test). The model's total explanatory power is substantial (conditional R^2=0.48). We also asked if the addition of potential confounding factors to the model is justified. Three factors were tested via the two above-mentioned methods: sex, brain bank site, and the anatomical location of the samples. In all cases, the AIC increased, and the P values by likelihood ratio tests were higher than 0.99. Therefore, from a statistical standpoint, the inclusion of these potential confounding factors does not seem to improve overall model fitting.

Minor corrections to the text and figures:

The authors made a great effort to analyze various samples from one individual donor. One can get a bit confused by the sentence that "an average of 2.5 brains samples were analyzed for each donor". Maybe the authors could highlight more in the first paragraph of the results section and in Figure 1A, that there are multiple samples ("technical replicates") from one individual patient across different brain regions used.

We removed the ‘2.5’ sentence and rewrote the paragraph for clarity. Samples information’s are now displayed in Table S1.

In the method section is a part included "Expression of target genes in microglia", it was very hard to allocate where these data from public data sets were actually used and for which analysis. Maybe the authors could clarify this again.

AU response: we apologize and corrected the paragraph in the methods (page 6) as follow: “ Expression of target genes in microglia. To evaluate the expression levels of the genes identified in this study as target of somatic variants, we consulted a publicly available database (https://www.proteinatlas.org/), and also plotted their expression as determined by RNAseq in 2 studies (Galatro et al. GSE99074 33, and Gosselin et al. 34) (Table S3 and Figure S2). For data from Galatro et al. (GSE99074) 33, normalized gene expression data and associated clinical information of isolated human microglia (N = 39) and whole brain (N = 16) from healthy controls were downloaded from GEO. For data from Gosselin et al. 34, raw gene expression ­data and associated clinical information of isolated microglia (N = 3) and whole brain (N = 1) from healthy controls were extracted from the original dataset. Raw counts were normalized using the DESeq2 package in R 35.”

Table S3 is very informative, but also very complex. The reader could maybe benefit a lot from this table if it can be structured a bit easier especially when it comes to identifying P-SNVs and in which tissue sample they were found and if this was the same patient. The sorting function on top of the columns helps, but the color coding is a bit unclear.

Despite our best efforts we agree that the table, which contain all sequencing data for all samples, is complex. The color coding (red) only highlights the presence of pathogenic mutation.

Reviewer #3 (Recommendations For The Authors):

This is a well-done study of an important problem. I present the following minor critiques:

At the bottom of Page 4 and into the top of Page 5, the authors state that 66 of the 826 variants identified in their panel sequencing experiment were found in multiple donors. Then the authors proceed to analyze the remaining 760 variants. It seems that the authors concluded that these multi-donor mosaics were artifacts, which is why they were excluded from further analysis. I think this is a reasonable assumption, but it should be stated explicitly so it is clear to the reader. Complicating this assumption, however, the authors later state that one of their CBL variants was found in two donors, and it is treated as a true mosaic. The authors should make it clear whether recurrent variants were filtered out of any given analysis. It remains possible that all recurrent variants are true mosaics that occurred in multiple donors. The authors should do a bit more to characterize these recurrent variants. Are they observed in the human population using a database like gnomAD, which, together with their recurrence, would strongly suggest they are germline variants? Are they in MAPK genes, or otherwise relevant to the study?

We apologize for the confusion. Our original intent for the ddPCR validation of variants (Figure 1E) was to count only 1 ‘unique’ variant for variants found for example in 1 brain sample and in the blood from the same patient, or in 2 brain regions from one patient, in order to avoid the criticism of overinflating our validation rate. This was notably the case for TET2 and DNMT3 variants. For example, validation of a TET2 variant found in 2 different brain areas and blood of the same donor is counted as 1 and not 3. We did not eliminate these variants from the analysis as they passed the criteria for somatic variants as presented in Methods.

In contrast, when a specific variant was found and validated in two different donors, we counted it as 2.

The characterization of variants included multiple parameters and databases, including for example AF and gnomAD, as indicated in Methods and reported in Table S3.

All ddPCR results can be found at the end of Table S3.

Figure 2B labels age-matched controls as "C", but Figure 2C labels age-matched controls as AM-C. Labels should be consistent throughout the manuscript.

We corrected this in the revised version.

It is not clear if the "p:0.02" label in Figure 2F is referring to AM-C Cx vs. AD-Cx or AM-C vs. AD. Please clarify.

We apologize for the confusion, and we corrected the legend. The calculated p value is for the comparison between Cortex from Controls (age-matched) and the Cortex from AD.

On Page 7, the authors state, "The allelic frequencies at which MAPK activating variants are detected in brain samples from AD patients range from ~1-6% of microglia (Fig. 3G), which correspond to clones representing 2 to 12% of mutant microglia in these samples, assuming heterozygosity." I understand what the authors mean here but I think it's a bit confusingly stated. I suggest something like "The allelic frequencies at which MAPK activating variants are detected in brain samples from AD patients range from ~1-6% in microglia (Figure 3G), which correspond to mutant clones representing 2 to 12% of all microglia in these samples, assuming heterozygosity."

We thank the reviewer for this suggestion and re-wrote that sentence.

Is there any evidence that the transcriptional regulators mutated in AD microglia (MED12, SETD2, MLL3, DNMT3A, ASXL1, etc.) are involved in regulating MAPK genes? This would tie these mutations into the broader conclusions of the paper.

This is a very interesting question, and indeed published studies indicate that some of the transcriptional /epigenetic regulators regulate expression of MAPK genes. However, in the absence of experimental evidence in microglia and patients, the argument may be too speculative to be included.

Do the authors have any thoughts as to whether germline variants in CBL are linked to AD? If not, why do they think germline mutations in CBL are not relevant to AD?

This is also a very interesting question. As indicated in our manuscript, germline mutations in CBL (and other member of the classical MAPK genes, see Figure 3C) cause early onset (pediatric) and severe developmental diseases known as RASopathies, characterized by multiple developmental defects, and associated with frequent neurological and cognitive deficits.

It is possible that some other (and more frequent?) germline variants may be associated with a late-onset brain restricted phenotype, but we did not find germline pSNV in our patients. GWAS studies may be more appropriate to test this hypothesis.

Do any donors show multiple variants? I don't think this is addressed in the text.

We do find donors with multiple variants (see Figure 3D and Figure S3), however at this stage, we did not perform single nuclei genotyping to investigate whether they are part of the same clone.

Figure S3 appears to be upside down.

This was corrected

Figure 5C should have some kind of label telling the reader what gene set is being depicted.

We added this information above the panel (it was in the corresponding legend).

At the top of Page 12, Lewy bodies are written as Lewis bodies.

This was corrected

Many control donors died of cancer (Table S1). Is there any information on which, if any, chemotherapeutics or radiation these patients received? Might this impact the somatic mutation burden? The authors should compare controls with and without cancer or with and without cancer treatments to rule this out.

As suggested by the reviewer, we analyzed the mutational load of age-matched controls with and without cancer (revised Figure S2B). As expected, we saw an increase in the mutational load in controls with cancer, particularly in their blood. This information was added in the result section.

This is most likely associated with the treatments received as well as possible cancer clones.

The formatting for Table S3 is odd. Multiple different fonts are used (this is also seen in Table S5). Column Q has no column ID. The word "panel" is spelled "pannel." The word "expressed" is spelled "expressd" in one of the worksheet labels. Columns BG-BN in the ALL-SNV worksheet are blank but seemingly part of the table.

We fixed this error in Table S3.

Version 2

Abstract

Summary

Somatic genetic heterogeneity resulting from post-zygotic DNA mutations is widespread in human tissues and can cause diseases, however few studies have investigated its role in neurodegenerative processes such as Alzheimer’s Disease (AD). Here we report the selective enrichment of microglia clones carrying pathogenic variants, that are not present in neuronal, glia/stromal cells, or blood, from patients with AD in comparison to age-matched controls. Notably, microglia-specific AD-associated variants preferentially target the MAPK pathway, including recurrent CBL ring-domain mutations. These variants activate ERK and drive a microglia transcriptional program characterized by a strong neuro-inflammatory response, both in vitro and in patients. Although the natural history of AD-associated microglial clones is difficult to establish in human, microglial expression of a MAPK pathway activating variant was previously shown to cause neurodegeneration in mice, suggesting that AD-associated neuroinflammatory microglial clones may contribute to the neurodegenerative process in patients.

Introduction

Neurodegenerative diseases are a frequent cause of progressive dementia. Alzheimer’s disease (AD) is diagnosed in ∼90% of cases, with an estimated prevalence of ∼10% in the population over 65 years of age 1,2. The role of germline genetic variation in neurodegenerative diseases and AD has been studied intensely. Although autosomal dominant forms of AD due to rare germline variants with high penetrance account only for an estimated ∼1% of cases 3–8, a number of common variants were also shown to contribute to disease risk. Carriers of one germline copy of the epsilon4 (E4) allele of the apolipoprotein E gene (APOE4), present in ∼15 to 20% of population, have a three-fold higher risk of AD, while two copies (∼2 to 3 % of population) increase the risk by ∼10-fold 9–12. Genome-wide association studies (GWAS) have identified an additional ∼50 common germline variants that more moderately increase the risk of AD, including TREM2, CD33, and MS4A6A variants 13–15. Interestingly, the APOE4 allele is responsible for an increased inflammatory and neurotoxic response of microglia and astrocytes in the brain of carriers 16–18, and it was noted that the majority of the other germline AD-risk variants are located within or near genes expressed in microglia 15 and in particular at microglia-specific enhancers 19. These data, together with transcriptional studies 20–22 support the hypothesis that genetic variation in microglia may contribute to the pathogenesis of neurodegeneration and AD.

Somatic genetic heterogeneity (mosaicism), resulting from post-zygotic DNA mutations, is widespread in human tissues, and a cause of tumoral, developmental, and immune diseases 23–26. Aditionally, a role of somatic variants in neuropsychiatric disorders is also suspected 27. Mosaicism has been documented in the brain tissue of AD patients in several deep-sequencing studies 28–30, showing that the enrichment of putative pathogenic somatic mutations in the PI3K-AKT, MAPK, and AMPK pathway do occur in the brain of patients in comparison to controls 30. However these studies performed in whole brain tissue lacked cellular resolution and mechanistic insights, and the role of somatic mutants in neurodegenerative diseases remains poorly understood 23. Somatic variants that activate the PI3K-AKT-mTOR or MAPK pathways in neural progenitors are a cause of cortical dysplasia and epilepsy 31–34 and developmental brain malformations 35, while somatic variants that activate the MAPK pathway in brain endothelial cells are associated with arteriovenous malformations 36. Interestingly, we reported that expression of a somatic variant activating the MAPK pathway in microglia causes neurodegeneration in mice 37, but the presence and contribution of microglial somatic clones in neurodegenerative diseases and AD remains unknown.

Here, we investigated the presence and nature of somatic variants in brain cells from control and AD patients. In an attempt to examine all brain cells at the same resolution, nuclei from neurons, glia cells and microglia, which only represent ∼5% of brain cells, were pre-sorted. Human microglia are reported to develop in embryo and renew by local proliferation within the brain 38–40. However, bone marrow-derived myeloid cells can enter the brain, in particular during pathological processes, and may not be distinguishable from resident microglia by transcriptomics alone 41. In order to distinguish somatic variants carried by resident microglia from the one carried by myeloid cells of peripheral origin, we analyzed matched peripheral blood from control and patients, to ‘barcode’ somatic mutants shared between microglia and blood. Finally, in order to achieve high sensitivity in the detection of variants that confer a proliferative or activation advantage (pathogenic mutations) and support the emergence or pathogenicity of mosaic clones 42, and/or that have been previously associated with neurological diseases, we initially performed a targeted deep-sequencing of a panel of 716 genes covering somatic variants reported in clonal proliferative disorders and genes associated with neurodegenerative diseases diseases.

We found that microglia from AD patients were enriched for pathogenic variants in comparison to age-matched controls. Furthemore, we found that these microglia-specific AD-associated variants preferentially target the MAPK pathway, including recurrent CBL ring-domain mutations. In addition, we showed that these variants drive a microglia transcriptional program characterized by a strong neuro-inflammatory response previously associated with neurotoxicity, including the production of IL1 and TNF, both in in vitro microglia models and in patients. The natural history of the AD-associated microglia clonal inflammatory disorder we describe here is difficult to establish. Specifically, we do not know whether it contributes to the onset of the neuro-inflammatory process at an early stage of the disease, or if microglia carrying pathogenic mutations preferentially expand later during the course of the disease in response to tissue inflammation. Under both hypotheses however, the presence of neuro-inflammatory microglial clones may contribute to the neurodegenerative process in a subset of AD patients. This report reveals a previously unrecognized presence of AD-associated microglia harboring pathogenic somatic variants in humans and provides mechanistic insight for neurodegenerative diseases by delineating cell-type specific variant recurrence.

Results

Clonal diversity among brain cells and blood from controls and AD patients

We examined post-mortem frozen brain samples and matching blood from 45 patients with intermediate-onset sporadic AD and 44 control individuals who died of other causes, including 27 donors age and sex-matched donors with the AD cohort (Fig. 1A; Supplementary Fig. 1A and Supplementary Table 1). APOE risk allele frequency for patients and controls was comparable to published series 10–12 (Supplementary Fig. 1A), and analysis of germline mutations did not identify deleterious variants in the 140 genes associated with neurological diseases. Myeloid/microglia, neurons, and glia/stromal cells were purified by flow cytometry using antibody against PU.1 and NeuN 43 (Fig. 1B, Supplementary Fig. 1B and 1C). Single nuclei (sn)RNAseq was performed on PU.1+ nuclei from one control and 3 AD patients to evaluate microglia enrichment following PU.1+ purification, and a cell-type annotation analysis indicated that ∼94% of PU.1+ nuclei correspond to microglia (Fig. 1C and Supplementary Fig. 1D-1H). Cortex samples were obtained from all donors but hippocampus samples were mostly obtained from AD patients (Fig. 1A; Supplementary Table 1). A total of 744 DNA samples from blood, PU.1+ nuclei, NeuN+ nuclei, and Double Negative nuclei (glia/stromal cells) from patients and controls (Fig. 1A) were submitted to targeted hybridization/capture and deep-DNA targeted sequencing (TDS, Fig. 1D, see Methods), at mean coverage of ∼1,100x (Supplementary Fig. 1I), for a panel of 716 genes (3.43 Mb, referred to below as BRAIN-PACT) which included genes reported to carry somatic variants in clonal proliferative disorders (n=576 genes)44,45 or that have been reported to be associated with neurodegenerativediseases (n=140) 46–54 (Supplementary Table 2, see Methods).

Detection of mutations in brain cell types and blood

(A) Table with patient and sample information. (B) Schematic represents the isolation and labelling of nuclei from post-mortem frozen brain samples from controls and Alzheimer’s disease patients with DAPI and antibodies against PU.1+ (myeloid/microglia) and NeuN+ (neurons).Representative flow cytometry dot-plot of nuclei separation. Double negative nuclei are labeled ‘DN’. (C) Percentage of cell types obtained in sorted PU.1+ nuclei determined by single-nuclei RNAseq in 5 brain samples from 4 individuals. (D) Schematic represents the sequencing strategy. Two algorithms (ShearwaterML and Mutect1) were used for variant calling. After annotation, pathogeneity was determined using OncoKb and ClinVar. (E) Venn diagram represents the number of variants and overlap between the ShearwaterML and Mutect1. Numbers in red indicate pathogenic variants (P-SNV). Validation of variants was performed by droplet digital (dd)PCR on pre-amplified DNA when available. (F) Venn diagrams represent the repartition per cell type of the 826 single-nucleotide variations (SNVs) identified in NeuN+: Neurons, PU.1+: microglia, DN: glia, and matching blood. [Numbers] in red indicate pathogenic variants P-SNV

After QC and filtering of germline variants, variant calling using ShearwaterML and a curated Mutect1 analysis identified 826 somatic synonymous and non-synonymous single-nucleotide-variations (SNVs), at an allelic frequency > 0.3% (mean 1.3%) in the 744 samples, corresponding to an overall variant burden of 0.3 mut/Mb (Fig. 1E). Sixty-six/826 SNV were present in more than one sample (Supplementary Table 3). Droplet digital-PCR performed on pre-amplification DNA for ∼10% of the 760 unique SNV was positive in 90% of cases (Fig.1E; Supplementary Table 3). After annotation using the OncoKB 55 and ClinVar 56 databases for disease-associated or causative variants (Fig. 2D,F; Supplementary Table 3), 96 unique SNV were classified as Pathogenic (P)-SNV. 40% of these P-SNV were tested by droplet digital-PCR and confirmed in 95% of cases (Fig. 1E and Supplementary Table 3). Positive and negative results in matching brain samples from individual donors were confirmed in 100% of samples at a mean depth of ∼5000x (range 648-23.000x) (Supplementary Table 3). A venn-diagram analysis of SNVs detected in PU.1+, NeuN+ , DN, and blood samples indicated that most (>90%) SNV and P-SNV were cell-type or tissue specific, with ∼ 5% of SNV and ∼ 8% of P-SNV shared between the blood and brain of individual donors (Fig. 1F; Supplementary Table 3). These data indicate that targeted deep-sequencing of purified nuclei allows to detect clonal mosaic variants with high sensitivity and specificity. In addition, ‘bar-coding’ of clonal variants across tissues suggests that infiltrating myeloid cells of peripheral origin account for ∼5% of microglia somatic diversity, and therefore that blood clones have a detectable but minor contribution to microglia, consistent with its local maintainance and proliferation 38,39.

Pathogenic variants are enriched in microglia from AD patients.

(A) Correlation plot represents the mean number of variants per cell type and donor (n=89) (Y axis), as a function of age (X axis). Each dot represents mean value for a donor. Statistics: fitted lines, the correlation coefficients (rs) and associated p values were obtained by linear regression (Spearman’s correlation). (B) Number of SNV per Mb and cell types per donor, of age-matched controls (n=27) and AD patients (n=45). Each dot represents mean value for a donor. Statistics: p-values are calculated with unpaired two-tailed Mann-Whitney U. Note: non-parametric tests were used when data did not follow a normal distribution (D’Agostino-Pearson normality test). (C) Number of SNV per Mb in PU.1+ samples across brain regions, of age-matched controls (n=27) and AD patients (n=45). Each dot represents a sample. Statistics: p-values are calculated with Kruskal–Wallis, multiple comparisons. Note: non-parametric tests were used when data did not follow a normal distribution (D’Agostino-Pearson normality test). (D) Correlation plot represents the mean number of pathogenic variants (P-SNV) as determined by ClinVar and/or OncoKB, per cell type and donor (n=89) (Y axis), as a function of age (X axis). Each dot represents mean value for a donor. Statistics: fitted lines, the correlation coefficients (rs) and associated p values were obtained by linear regression (Spearman’s correlation). (E) Number of P-SNV per Mb and cell types per sample, of age-matched controls (n=27) and AD patients (n=45). Each dot represents a sample. Statistics: p-values are calculated with unpaired two-tailed Mann-Whitney U. Note: non-parametric tests were used when data did not follow a normal distribution (D’Agostino-Pearson normality test). (F) Number of P-SNV per Mb in PU.1 samples across brain regions, of age-matched controls (n=27) and AD patients (n=45). Each dot represents a sample. Statistics: p-values for comparison within each group (controls and patients) are calculated with Kruskal–Wallis test and Dunn’s test for multiple comparisons. p-values for the comparison of P-SNV between the cortex of C and the cortex of AD (0.01) was calculated with unpaired two-tailed Mann-Whitney U. Note: non-parametric tests were used when data did not follow a normal distribution (D’Agostino-Pearson normality test). (G) Number of P-SNV per Mb and and cell types per donor for age-matched controls (n=27) and AD patients (n=45). Each dot represents mean value for a donor. Statistics: p-values are calculated with unpaired two-tailed Mann-Whitney U test. Odds ratio (95% CI, 2.049 to 29.02) and p values for the association between AD and the presence of pathogenic variants are calculated by multivariate logistic regression, with age and sex as covariates. (H) P-SNV burden as a function of age and disease status (age-matched controls and AD). Linear lines represent trend lines from mixed-effects linear regression that incorporates individual donor as a random effect (blue, control: P=0.0025, R^2=0.13; red, NDD: P=9.1x10^-16, R^2=0.50 by Pearson’s correlation). The model’s total explanatory power is substantial (conditional R^2=0.48). Both age and AD are associated with a significant increase in SNV burden in this model (P<1x10^-4 and P=1x10^-4, respectively, by likelihood ratio test). Anatomical regions of the brain specimen and originating brain banks were not incorporated because the models incorporating those parameters did not significantly improve the overall model fitting by likelihood ratio test (see Methods). Graph depicts SNV burden corrected by the mixed-effects model (See Fig S2D for observed P-SNV burden)

Somatic clonal diversity of the different cell types, as evaluated by the SNV/megabase burden was higher in blood (1 mut/Mb) and PU.1+ nuclei (0.5 mut/Mb) than for DN and neurons (0.18 mut/Mb) (Supplementary Fig. 2A). The SNV/mb burden of blood and PU.1+ nuclei increased as a function of age (Fig. 2A and Supplementary Table 3) as previously reported for proliferating cells 24,57–59. Interestingly, the SNV/mb burden of blood cells from age-matched controls was higher than for AD patients (Fig. 2B). In contrast, there was no difference in SNV/mb burden between PU.1+, NEUN+ and DN samples from AD patients and age-matched controls (Fig. 2B), and between PU.1+ nuclei from the cortex, hippocampus, and brainstem/cerebellum samples (Fig. 2C). These data altogether indicate that the clonal diversity of microglia and blood both increase with age, and that the clonal diversity of blood cells is lower in AD than in age-matched controls who died of other causes including cancer and cardiovascular diseases (see Methods). This is consistent with recent studies showing that clonal hematopoiesis is associated with a higher risk of several diseases related to ageing such as cardiovascular diseases, but inversely associated with the risk of AD 60,61.

Microglia clones carrying pathogenic variants are enriched in AD patients

In contrast to the global SNV burden, increased P-SNV burden was correlated not only with age (Fig. 2D), but also with the disease status (AD) (Fig. 2E-H). Within the control group, the SNV and P-SNV burden was higher in the blood of controls treated for cancer (Supplementary Fig. 2B). The P-SNV burden per Mb was selectively and highly enriched in PU.1+ samples from AD patients in comparison to age-matched controls (p=0.0003, Fig. 2E). Analysis of PU1+ P-SNV/Mb burden per brain region indicated that the P-SNV/Mb burden was similar between brain regions within each group (Fig. 2F and Supplementary Fig. 2C), and therefore attributable to AD status rather than sampling bias. Analysis of mutational load per donor confirmed that microglial clones carrying P-SNV were enriched in the brain of AD patients in comparison to age-matched controls (Fig. 2G). Despite the relatively modest cohort size, a logistic regression analysis confirmed the association between the presence of P-SNVs in PU.1+ nuclei and AD after adjusting for sex and age (OR= 7; p=0.0035, Fig. 2G and Supplementary Fig. 2D). A mixed-effects linear regression model analysis also showed an excess of P-SNVs in AD independently of the effect of age (p = 0.0215) (Fig. 2H and Supplementary Fig. 2E).

In addition, genes targeted by P-SNV were all expressed in microglia (Supplementary Fig. 2F and Supplementary Table 3) and the analysis of P-SNV/Mb mutational load restricted to genes that are not expressed in microglia did not show an enrichment of candidate pathogenic variants in AD patients (Supplementary Fig. 2G and Supplementary Table 3). Altogether, these results show an association between microglia clones carrying P-SNV and AD in this series.

AD patients carry microglial clones with MAP-Kinase pathway variants including recurrent CBL variants

Pathways analysis of genes carrying P-SNV in microglia from AD patients, against the background of the 716 genes sequenced, showed that the most significant pathways enriched were the receptor tyrosine kinase /MAP-Kinase pathways (Reactome, GO, and canonical pathways, Fig. 3A and Supplementary Table 4), corresponding to pathogenic/oncogenic variants in 6 of the 15 genes of the classical MAPK pathway 62 (CBL, BRAF, RIT1, NF1, PTPN11, KRAS), TEK, and the KEGG Chronic Myeloid Leukemia (CML) pathway, which includes the former plus SMAD5 and TP53 (Fig. 3B, 3C and Supplementary Fig. 3). Mutational load for MAPK genes was significantly higher in AD patients in comparison to age-matched control (Fig. 3C). Other enriched pathways, albeit less significant, included genes involved in DNA repair and chromatin binding/ methyltransferase activity (Fig. 3B; Supplementary Table 4). No pathway was enriched in age matched controls. Of note we did not observe microglia P-SNVs within genes reported to be associated with neurological disorders (Supplementary Table 2) in patients (Supplementary Table 3). P-SNV targeting genes of the classical RTK/ MAPK pathway (Fig. 3C) were detected in the PU.1+ samples from ∼25% of the AD patients tested (p=0.0145 vs age-matched controls, Fig. 3D and Supplementary Fig. 3). Strikingly, half of these patients (6 patients, 13% of AD patients in this series) carried reccurent P-SNV in the RING domain of CBL 63–73 (Fig. 3B-3E). Two additional patients presented with P-SNV in the Switch II domain of RIT1 74 (Fig. 3B-3F). Microglia from the 3 other patients carried activating KRAS (p.A59G), PTPN11 (p.T73I) and TEK (p.R1099*) oncogenic variants previously described in cancer and sporadic venous malformations 75–77 (Fig. 3B and 3D). In addition, a 12th patient carried a gain of function (GOF) U2AF1 (p.S34F) variant 78, which is not a ‘classical MAPK gene’ but activates the MAPK pathway in myeloid malignancies 79 (Fig. 3B and Supplementary Fig. 3). Two patients carried 2 different MAPK activating variants: microglia from 1 patient carried an activating BRAF (p.L505H) variant 80 in addition to loss of function (LOF) variant CBL (p.C416S), and another patient carried the NF1 (p.L2442*) LOF variant 81,82 in addition to the activating RIT1 (p.M90I) variant (Fig. 3D and Supplementary Fig. 3). Five patients also carried additional P-SNV targeting genes involved in DNA repair with tumor suppressor function 83,84, including the loss of function variants in ATR (c.6318A>G) 85 and SMC1A (p.X285_splice) (Fig. 3D and Supplementary Fig. 3), and in DNA/histone methylation including TET2 (p.Q1627*) 63,86, IDH2 (p.R140Q) 87, and PBRM1 (c.996-7T>A) 88) (Fig. 3D and Supplementary Fig. 3). Finally, two patients carried oncogenic variants in genes from the KEGG Chronic Myeloid Leukemia (CML) pathway, SMAD3 (p.R373C) 89 and TP53 (pX261_splice) 90 (Fig. 3D and Supplementary Fig. 3). The detection of multiple oncogenic variants in the same patients is reminiscent of the features observed in myeloproliferative disorders described outside the brain 73,86.

Somatic microglial clones with multiple and recurrent CBL and MAP-Kinase pathway activating variants.

(A) Pathway enrichment analysis for the genes target of P-SNVs using the panel of 716 genes as background set. Graph shows the most enriched pathways by: Reactome Gene Sets, GO Molecular Functions, Canonical Pathways and KEGG Pathway (see complete list in Supplementary Table 4). (B) Bar plot indicates the genes carrying P-SNV (y-axis) and the % of AD patients carrying P-SNV for each gene (x-axis). Genes are color-coded by pathway. (C) Representation of the classical MAPK pathway, the 6 genes mutated in AD patients are labeled in red, TEK is labeled in blue, and larger font size indicate reccurence of variants in a given gene. Violin plot shows distribution of P-SNV in genes from classical MAPK pathway per Mb sequenced and per sample in patients and controls, p-value: unpaired two-tailed Mann-Whitney U test. (D) Summary Table showing patients carrying P-SNV in the classical RTK/MAPK pathway and CML associated genes (see Supplementary Table 3) and indicating the detection of variants in blood, and their association with other variants in microglia. (E) Recurrent variants in the ring-like domain of CBL are indicated in red on the diagram structure of gene, above representative Western blot from cell lysates from HEK293T cells expressing WT, positive control (Y371H), or CBL variants alleles found in patients, and stimulated with EGF or control, probed with antibodies against Phospho-p44/42 MAPK (Erk 1/2, Thr202/Tyr204), total p44/42 MAPK (Erk1/2), HA-tag and tubulin (BOTTOM). Histogram (RIGHT) represents quantification of the increase of the Phospho-ERK1/2 /total ERK1/2 ratio in western blots in n=5 independent experiments, statistics: unpaired one-tailed t-test. (F) RIT1 M90I and F82L are represented on the 3D structure of the gene (pdb code: 4klz, F82 is within a segment whose structure was not resolved) and representative western blot from HEK293T cells expressing Flag-RIT1 (WT and mutants) and treated -/+ 20% FBS before harvesting. Lysates were probed with antibodies against Phospho-p44/42 MAPK (Erk 1/2, Thr202/Tyr204), total p44/42 MAPK (Erk1/2, (MAPK)), Flag, and tubulin. Histogram (RIGHT) represents quantification of the increase of the Phospho-ERK1/2 /total ERK1/2 ratio in western blots in n=4 independent experiments, statistics: unpaired one-tailed t-test. (G) Variant allelic frequency (VAF, %) for the BRAFV600E allele in PU.1+ nuclei from brain samples from histiocytosis patients (each dot represents a sample) and for P-SNVs in in PU.1+ nuclei from brain of AD patients (each dot represent a variant). Note: non-parametric tests were used when data did not follow a normal distribution (D’Agostino-Pearson normality test). (H) Percentage of P-SNVs detected by targeted deep sequencing (TDS) which were also detected by Whole-Exome-Sequencing (WES).

Recurrent CBL and RIT1 variants activate the MAPK pathway

CBL is an E3 ubiquitin-protein ligase that negatively regulates RTK signaling via MAPK 91. CBL somatic and germ-line LOF variants such as R420Q have been previously associated with tumoral diseases including clonal myeloproliferative disorders 63–73 and RASopathies 92 respectively. We confirmed that CBL RING-domain variants found in AD patients increased MAPK phosphorylation in response to EGF upon expression of HA-tagged WT or mutant alleles in HEK293T cells (Fig. 3E and Supplementary Fig. 4A). RIT1 is a RAS GTPase, and somatic or germ-line GOF variants such as RIT1 F82L and RIT1 M90I, also enhance MAPK signaling in malignancies 74 and RASopathies 93,94 respectively. As in the case of CBL variants, the 2 RIT1 variants found in AD patients increased MAPK phosphorylation in response to FBS in HEK293T cells expressing these mutant alleles (Fig. 3F, Supplementary Fig. 4B and 4C). These data altogether indicate that a subset of AD patients (12/45, ∼ 27% of this series) present with microglial clones carrying one or several oncogenic variants that activate the RTK/MAPK pathway, and are characterized by recurrent oncogenic variants in CBL and RIT1.

Allelic frequency of the patients’ MAPK activating variants

The allelic frequencies at which MAPK activating variants are detected in brain samples from AD patients range from ∼1-6% in microglia (Fig. 3G), which correspond to mutant clones representing 2 to 12% of all microglia in these samples, assuming heterozygosity. This range of allelic frequency is frequently observed for the MAPK activating BRAFV600E variant in microglia isolated from brain samples of 6 patients diagnosed with BRAFV600E+ histiocytosis, a rare clonal myeloid disorder associated with neurodegeneration 37,95–98 (Fig. 3G and Supplementary Table 5). These data suggested that the size of the mutant microglial clones in AD patients was compatible with a role in a neuro-inflammatory/neurodegeneration process.

Other variants found in microglia from AD patients

Pathogenic variants that did not involve the MAPK pathway included LOF variants in the DNA repair gene CHEK2 including CHEK2 c.319+1G>A 99 and CHEK2 R346H (Supplementary Fig. 3 and 4D), Mediator Complex gene MED12 100, Histone methyltransferases SETD2 101 and KMT2C/MLL3, the DNA methyltransferase DNMT3A 86,102, DNA demethylating enzymes TET2 and the Polycomb proteins ASXL1 86. Of note, TET2, DNMT3 and KMT2C variants when present, were frequently detectable in the patients’ matching blood at low allelic frequency (Supplementary Fig. 3). TET2, DNMT3 and KMT2C are frequently mutated in clonal hematopoiesis 58,59, suggesting that in contrast to other variants, the presence of TET2, DNMT3 and KMT2C/MLL3 in the brain of patients may reflect the entry of blood clones in the brain.

In half of the AD patients, no microglia pathogenic variants were identified. Targeted deep sequencing (TDS) cannot identify variants located outside of the BRAIN-PACT panel, such as other potential additional variants that would activate the MAPK pathway. Therefore, we performed whole exome sequencing (WES) of PU.1+ nuclei at an average depth ∼400x, in selected samples from 48 donors, including samples from most of the patients negative for pathogenic variants by TDS (n=17 out of 22), a selection of patients with variants identified by TDS (n=16 out of 23), and 15 controls, followed by a curated Mutect analysis. Only 6/15 (40%) of the pathogenic SNVs previously identified by TDS and confirmed by ddPCR were detectable by WES in these samples (Fig. 3H), indicating a lower sensitivity of WES. Nevertheless, after annotation by 4 modeling predictors (Polyphen, SIFT, CADD/MSC and FATHMM-XF 103–108 additional SNVs predicted to be deleterious with high confidence were identified in 8/22 patients without pathogenic variants identified by TDS (Supplementary Fig. 3 and Supplementary Table 6). Interestingly, 4 of the predicted deleterious variants identified by WES targeted genes that regulate the MAPK pathway (ARHGAP9, ARHGEF26, CHD8, and DIXDC1 (Supplementary Fig. 3 and SupplementaryTable 6).

The patients’ MAPK activating variants increases ERK phosphorylation, proliferation, inflammatory and mTOR pathways in murine microglia and macrophages

CBL variants increased ERK phosphorylation upon lentiviral transduction in BV2 murine microglial cells 109,110 (Supplementary Fig. 4E). However as this line was immortalized by v-Raf, which might interfere with the study of the MAPK pathway, we also stably expressed WT and variant CBL, RIT1, KRAS, PTPN11 alleles in SV-U19–5 transformed mouse ‘MAC’ lines 111,112 (see Methods and Supplementary Fig. 5A,B). MAC lines expressing CBL, RIT1, KRAS and PTPN11 variants presented with increased ERK phosphorylation and/or increased proliferation in comparison to their WT controls, as measured by Western immunoblotting and EdU incorporation (Fig. 4A and Supplementary Fig. 5A,B). In addition, Hallmark and KEGG pathway analysis of RNAseq data from control and mutant lines showed increased RAS, TNF, IL6 and JAK STAT signaling, complement, inflammatory responses, and mTOR pathway activation signatures in mutants (Fig. 4B and Supplementary Table 7). These data indicated that microglia variants from patient’s activate murine microglial cells and growth factor-dependent macrophages with proliferative and inflammatory responses in vitro. However, overexpression of mutant alleles in mouse cell lines does not necessarily recapitulate or predict the effects of a heterozygous genetic variant in physiological conditions. Thus, we investigated the role of CBLC404Y allele in heterozygous human primary microglia-like cells.

MAPK pathway activating variants in mouse macrophages and human iPSC-derived microglia-like cells.

(A) Representative western-blot analysis (Top panels) and quantification (Middle panels) of phospho-and total-ERK in lysates from a murine CSF-1 dependent macrophage cell line expressing CBLWT, CBLI383M, CBLC384Y, CBLC404Y, CBLC416S (n=3-6), and RIT1WT, RIT1F82L and RIT1M90I (n=3), KRASWT and KRASA59G (n=3). Bottom panels depicts flow cytometry analysis of EdU incorporation in the same lines. Statistics, Unpaired t-test. (B) HALLMARK and KEGG pathways (FDR / adj.p value <0.25, selected from Supplementary Table 7) enriched in gene set enrichment analysis (GSEA) of RNAseq from mutant CSF-1 dependent macrophages lines CBLI383M, CBLC384Y, CBLC404Y, CBLC416S, CBLR420Q, RIT1F82L RIT1M90I, KRASA59G, and PTPN11T73I (n=3-6) in comparison with their wt controls. NES: normalized enrichment score. (C) Sanger sequencing of 2 independent hiPSC clones (#93 and #91) of CBL404C/Y heterozygous mutant carrying the c.1211G/A transition on one allele and 2 independent isogenic control CBL404C/C clones (#71 and #89) all obtained by prime editing. (D) Photomicrographs in CBL404C/C and CBL404C/Y iPSC-derived microglia-like cells.(E) Quantification of leading edge and lateral lamellipodia in CBL404C/C and CBL404C/Y iPSC-derived microglia-like cells. n=3-7, statistics: p-value are obtained by nested one-way ANOVA. (F) Flow cytometry analysis of cell size for the same lines (n>3) statistics: p-value are obtained with nested one-way ANOVA,). (G) Flow cytometry analysis of EdU incorporation in CBL404C/C and CBL404C/Y microglia-like cells after a 2 hours EdU pulse. n=3, unpaired t-test). (H) Western-blot analysis (left) and quantification (right) of phospho- and total-ERK proteins in lysates from CBL404C/C and CBL404C/Y microglia-like cells starved of CSF-1 for 4 h and stimulated with CSF-1 (5 min, 100 ng/mL) (n=4), statistics: pvalue are obtained with two-way ANOVA.

Heterozygosity for a CBL variant allele activates human microglia-like cells

We used prime editing 113 of human induced pluripotent stem cells (hiPSCs, see Methods) to generate isogenic hiPSCs clones heterozygous for the patients’ variants (Fig. 3C and Supplementary Fig. 5). We focused our analysis on CBL404C/Y mutant lines because CBL was mutated in 6 patients and 2 of them carried the same CBL c.1211G>A p.C404Y variant (Fig. 3D). Microglia-like cells were differentiated from two independent hiPSC-derived CBL404C/Y lines and their isogenic CBL404C/C controls (Fig. 4C and Supplementary Fig. 5C). CBL404C/Y and isogenic CBL404C/C microglia-like cells expressed similar amount of CBL total mRNA and protein, and CBL404C/Y cells expressed wt and mutant mRNA in similar amounts, as expected assuming bi-allelic expression of CBL (Supplementary Fig. 5D-5F). CBL404C/Y cells presented with a phenotype comparable to isogenic CBL404C/C microglia-like cells for expression of IBA1, CSF1R, NGFR, EGFR, CD11b, MRC1, CD36, CD11c, Tim4, CD45, and MHC Class II (Supplementary Fig. 5G). Their viability was also comparable to control (Supplementary Fig. 5H). However, CBL404C/Y cells were larger and presented with more lamellipodia, resulting in an amoeboid morphology less frequently observed in isogenic controls (Fig. 4D,4E), and their proliferation rate was slightly increased, as measured by EdU incorporation (Fig. 4F). Moreover CBL404C/Y cells cultured in CSF1-supplemented medium also presented with a higher basal pERK level than control when restimulated with CSF1 (Supplementary Fig. 5I), and ERK phosphorylation after stimulation of starved microglia-like cells with CSF-1 was increased by ∼2 fold in comparison to isogenic WT (Fig. 4G). Altogether, these results showed that heterozygosity for a CBLC404Y allele is sufficient to activate human microglia-like cells increasing their proliferation and ERK activation.

Heterozygosity for a CBLC404Y allele drives a microglial neuroinflammatory /AD associated signature

Gene Set Enrichment Analyses (GSEA) of RNAseq comparing CBL404C/Y and isogenic CBL404C/C microglia-like cells showed upregulation of Glycolysis, Oxidative Phosphorylation, and mTORC1 signatures, indicating increased metabolism and energy consumption by the mutant cells (Fig. 5A and Supplementary Table 8). In addition, as observed in MAC lines, CBL404C/Y cells upregulated complement, TNF, and JAK STAT signaling and inflammatory signatures (Fig. 5A; Supplementary Table 8) 114. Increased production of TNF, IL-6, IFN-ψ, IL-1β, C3 and complement Factor H (CFH) by CBL404C/Y cells was confirmed by ELISA (Fig. 5B). In addition, CBL404C/Y microglia-like cells also presented with signatures from the KEGG database associated with neurodegenerative disorders (Fig. 5A; Supplementary Table 8), and for the recently published human microglia AD scRNAseq signature, obtained by analysis of 24 sporadic AD patients and 24 controls 21 (Fig. 5C). These data indicated that heterozygosity for the CBLC404Y allele is sufficient to drive expression of a neuroinflammatory /AD signature in a human microglia-like cell type, characterized by increased metabolism and the production of neurotoxic cytokines known to interfere with normal brain homeostasis.

CBL404C/Y microglia signature.

(A) HALLMARK and KEGG pathways (FDR /adj.p value <0.25, selected from Supplementary Table 8) enriched in gene set enrichment analysis (GSEA) of RNAseq from from CBL404C/Y iPSC-derived macrophages and isogenic controls NES, normalized enrichment score. (B) ELISA for pro-inflammatory cytokines (n=3) and complement proteins (n=2) in the supernatant from CBL404C/Y iPSC-derived microglial like cells and isogenic controls. Statistics: p-value are obtained by nonparametric Mann-Whitney U test,* 0.05, ** 0.01, *** 0.001, **** 0.0001. (C) GSEA analysis for enrichment of the human AD-microglia snRNAseq signature (MIC1) 21 in differentially expressed genes between CBL404Y/C microglial like cells and isogenic controls. (D) Dot plot represents the GSEA analysis of HALLMARK and KEGG pathways enriched in snRNAseq microglia clusters (samples from all donors). Genes are pre-ranked per cluster using differential expression analysis with SCANPY and the Wilcoxon rank-sum method. Statistical analyses were performed using the fgseaMultilevel function in fgsea R package for HALLMARK and KEGG pathways. Selected gene-sets with p-value < 0.05 and adjusted p-value < 0.25 are visualized using ggpubr and ggplot2 R package (gene sets/pathways are selected from Supplementary Fig. 6B, Supplementary Table 9). (E) Dot plot represents the GSEA analysis (as in (E)) of HALLMARK and KEGG pathways enriched in cluster 2 / 2B and deconvoluted by donor samples (selected from Supplementary Fig. 6A).

The MAPK variant neuroinflammatory microglial signature is detectable in patients

Analysis of the snRNAseq data from 5 samples of purified microglia nuclei from 4 donors (control, AD without and with pathogenic variants (Fig.1C, Supplementary Fig. 1D-H and Supplementary Table 9) using unsupervised Louvain clustering and GSEA showed that microglia samples from patients carrying variants were enriched for the signatures observed in the MAC lines and CBL404C/Y cells (Fig. 5D,E and Supplementary Fig. 6). In particular microglia cluster 2 and 2B, were most enriched for the inflammatory, TNF, mTOR and oxidative phosphorylation and glycolysis signatures in patients carrying variants (AD52, AD53) but not the controls (C11, AD34) (Fig. 5D,E and Supplementary Fig. 6). Despite the small size of the mutant clones and the low sensitivity of scRNAseq to detect rare allelic variants, KRAS A59G variant reads were detected in cluster 2/2B from patient AD52.

Altogether, the above results support the hypothesis that patients’ microglial clones carrying pathogenic mutations are associated with a metabolic and neuroinflammatory signature that includes the production of neurotoxic cytokines in vitro and in vivo.

Discussion

We report here that microglia from a cohort of 45 AD patients with intermediate-onset sporadic AD (mean age 65 y.o) is enriched for clones carrying pathogenic/oncogenic variants in genes associated with clonal proliferative disorders (Supplementary Table 2) in comparison to 44 controls. Of note we did not observed microglia P-SNVs within genes reported to be associated with neurological disorders in the patients.

These pathogenic variants are absent from blood, glia or neurons in most cases. They are found predominantly in the MAPK pathway and include reccurent variants (CBL RING domain variants in 6 patients), which promote microglial proliferation, activation, and expression of a neuroinflammatory/neurodegereration-associated transcriptional programme in vitro and in vivo, and the production of neurotoxic cytokines IL1b, TNF, and IFNg 115–118. Heterozygous expression of pathogenic CBL variant in human microglia-like cells was sufficient to drive a transcriptional program that associates with increased metabolic activity and a neurotoxic inflammatory response, also observed in microglia from patients with MAPK-activating variants.

The association between AD and MAPK pathway variants is consistent with a previous study where WES performed on unseparated brain tissue from AD patients showed that putative pathogenic somatic variants were enriched for the MAPK pathway, despite the lower sensitivity of the approach and the lack of cellular specificity 30. The pathogenic role of the somatic pathogenic variants in the MAPK pathway associated with the microglia of AD patients is supported by several lines of evidence. We show here that they promote a neuroinflammatory/neurodegereration-associated transcriptional programme in microglia like cells. In addition, somatic variants that activate the MAPK pathway in tissue macrophages cause a clonal proliferative and inflammatory disease called Histiocytosis, strongly associated with neurodegeneration 37,95–97, and introduction in mouse microglia of the variant allele most frequently associated with histiocytosis (BRAFV600E) causes neurodegeneration in mice 37. The allelic frequencies of pathogenic variants found in AD patients is lower than values classically observed in solid tumors or leukemia, but within the range of the clonal frequency of pathogenic T cells observed in auto-immune diseases 119, and we found that they were in the range of the allelic frequencies observed for the BRAFV600E variant in microglia in the brain of Histiocytosis patients. Moreover, the RAS/MAPK signaling pathway is involved in microglia proliferation, activation and inflammatory response 120–122, neuronal death, neurodegeneration, and AD pathogenesis 15,19,37,123, and its activation has been proposed to be an early event in the pathophysiology of AD in human 124. Neuroinflammation is an early event in AD pathogenesis, increasingly considered as critical in pathogenesis initiation and progression 16,125,126. This is underscored by the observation that the main known genetic risk factor for sporadic AD is the APOE4 allele, responsible for an increased inflammatory response in the brain of APOE4 carriers 16. In this regard, the contributing role of MAPK activating variants could be comparable to that of the APOE4 allele, and we noted that the allelic frequency of APOE4 allele is lower in patients with pathogenic variants (16/46 alleles, 34%) than patients without detected variant (23/44 alleles, 53%) although the difference did not reach significance in this series.

Variants targeting the DNA-repair and DNA/histone methylation pathways are also enriched among AD patients, sometimes associated in the same patients, albeit their functional significance was not investigated here. Of note however, germline variants of the DNA-repair transcription factor TP53, and DNA damage sensors ATR and CHEK2 were shown to promote accelerated neurodegeneration in human 83,84.

Microglia variants are frequently absent from blood, and our DNA sequencing barcoding approach does not support a model where blood cells massively infiltrate the brain or replace the microglia pool in patients from our series, but instead consistent with the local maintenance and proliferation of microglia 38,39. In addition, our results are consistent with a recent study showing that clonal hematopoiesis was inversely associated with the risk of AD61.

The association of microglia clones carrying pathogenic variants with AD in a subset of patients is not a consequence of an overall increase in microglia mutational load (SNV) in AD. Together with evidence that pathogenic variants drive neuroinflammation, these data suggest that these clones could contribute to AD pathogenesis, together with other genetic and environmental factors. Lewy bodies, amyloid angiopathy, tauopathy, or alpha synucleinopathy, were equally distributed among AD patients with or without microglia clones carrying MAPK activating variants. The natural history of the microglial clones is difficult to study in human. It is possible that microglial clones with proliferative and activation advantage and a neuroinflammatory and neurotoxic profile may be present at the onset and contribute to the early stages of the disease. Alternatively it is also possible that the microglial clones carrying the pathogenic mutations appear or are selected later during the course of the disease in the inflammatory milieu of the AD brain. In the latter case, pathogenic microglial clones may contribute to disease progression, i.e. neuroinflammation and neurodegeneration.

Competing interests

FG has been a paid consultant (no equity) to Third Rock Ventures from 2018 to 2020. Sequencing costs and analysis in this study were covered in part by a SRA between Third Rock venture and MSKCC. This work led to patents PCT/US2022/037893/WO2023004054A1 ‘Methods and compositions for the treatment of alzheimer’s disease’ by MSKCC and PCT/US2018/047964 ‘Kinase mutation-associated neurodegenerative disorders by MSKCC’.